\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R:}\\ Beyond repeated measures ANOVA and Linear Mixed Models\\ \strut \\ SUPPLEMENTARY MATERIALS: APPENDIX

### 2021-10-13 Manuscripts/Manuscript by chapters- SIM Revisions/00-Full_document_No_Appendix.Rmd
Started changes in Appendix, need to figure how to break up long code chunks and add text in between.

---
title: '**Using generalized additive models to analyze biomedical non-linear longitudinal data**'
subtitle: _Beyond repeated measures ANOVA and Linear Mixed Models_
header-includes:
    \usepackage{placeins}
output:  
  bookdown::pdf_document2:
    #template: my-template.tex #if a custom template that removes the additional "and" in the author information is desired
    pandoc_args: --listings #calls the listings package to fit code within the page margins
    keep_tex: yes #keep LaTeX file for submission
    fig_caption: yes #allows captions in figures
    toc: false #do not include table of contents
    extra_dependencies:
      subfig: null #allows for subfigures
      breqn: null #line breaks for long equations
      caption: ["font={small}"] #size of the figure captions
      #float: null #allows for control of placement of figures
    includes:
      in_header: preamble_OUP.sty #additional LaTeX formatting
  bookdown::word_document2:
    fig_caption: yes #figure caption
    keep_md: yes
  bookdown::html_document2:
    css: "style.css" #style for the HTML document
csl: biostatistics.csl #style for references
bibliography: refs.bib #references
link-citations: yes #adds links to the citations
'': default
---

# Summary

>>>>

---
title: '**Using generalized additive models to analyze biomedical non-linear longitudinal data**'
subtitle: _Beyond repeated measures ANOVA and Linear Mixed Models_
header-includes:
    \usepackage{placeins}
output:  
  bookdown::pdf_document2:
    #template: my-template.tex #if a custom template that removes the additional "and" in the author information is desired
    fontsize: 12pt
    fontfamily: helvetica
    pandoc_args: --listings #calls the listings package to fit code within the page margins
    keep_tex: yes #keep LaTeX file for submission
    fig_caption: yes #allows captions in figures
    toc: false #do not include table of contents
    extra_dependencies:
      subfig: null #allows for subfigures
      breqn: null #line breaks for long equations
      caption: ["font={small}"] #size of the figure captions
      #float: null #allows for control of placement of figures
    includes:
      in_header: preamble.sty #additional LaTeX formatting
  bookdown::word_document2:
    fig_caption: yes #figure caption
    keep_md: yes
  bookdown::html_document2:
    css: "style.css" #style for the HTML document
csl: american-medical-association.csl #style for references
bibliography: refs.bib #references
link-citations: yes #adds links to the citations
'': default
---

# Abstract

### 2021-10-13 Manuscripts/Manuscript by chapters- SIM Revisions/00-Full_document_SIM_No_Appendix.Rmd
Started changes in Appendix, need to figure how to break up long code chunks and add text in between.

---
title: '**Using generalized additive models to analyze biomedical non-linear longitudinal data**'
subtitle: _Beyond repeated measures ANOVA and Linear Mixed Models_
header-includes:
    \usepackage{placeins}
output:  
  bookdown::pdf_document2:
    #template: my-template.tex #if a custom template that removes the additional "and" in the author information is desired
    fontsize: 12pt
    fontfamily: helvetica
    pandoc_args: --listings #calls the listings package to fit code within the page margins
    keep_tex: yes #keep LaTeX file for submission
    fig_caption: yes #allows captions in figures
    toc: false #do not include table of contents
    extra_dependencies:
      subfig: null #allows for subfigures
      breqn: null #line breaks for long equations
      caption: ["font={small}"] #size of the figure captions
      #float: null #allows for control of placement of figures
    includes:
      in_header: preamble.sty #additional LaTeX formatting
  bookdown::word_document2:
    fig_caption: yes #figure caption
    keep_md: yes
  bookdown::html_document2:
    css: "style.css" #style for the HTML document
csl: american-medical-association.csl #style for references
bibliography: refs.bib #references
link-citations: yes #adds links to the citations
'': default
---

@@

In contrast, _generalized additive models_ (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures. Therefore, GAMs present an excellent choice to analyze non-linear longitudinal data in the context of biomedical research. This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data. We also present the basic theory of GAMs, and using trends of oxygen saturation in tumors reported in the biomedical literature, we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate how these models can be computationally implemented. We show that GAMs are able to produce estimates that are consistent with the trends of biomedical non-linear data even in the case when missing observations exist (with 40% of the simulated observations missing), allowing reliable inference from the data. To make this work reproducible, the code and data used in this paper are available at: https://github.com/aimundo/GAMs-biomedical-research.

>>>>

---
title: '**Generalized additive models to analyze biomedical non-linear longitudinal data in R: Beyond repeated measures ANOVA and Linear Mixed Models**'
header-includes:
    \usepackage{placeins}
output:  
  bookdown::pdf_document2:
    #template: my-template.tex #if a custom template that removes the additional "and" in the author information is desired
    fontsize: 12pt
    fontfamily: helvetica
    pandoc_args: --listings #calls the listings package to fit code within the page margins
    keep_tex: yes #keep LaTeX file for submission
    fig_caption: yes #allows captions in figures
    toc: false #do not include table of contents
    extra_dependencies:
      subfig: null #allows for subfigures
      breqn: null #line breaks for long equations
      caption: ["font={small}"] #size of the figure captions
      #float: null #allows for control of placement of figures
    includes:
      in_header: preamble.sty #additional LaTeX formatting
  bookdown::word_document2:
    fig_caption: yes #figure caption
    keep_md: yes
  bookdown::html_document2:
    css: "style.css" #style for the HTML document
csl: american-medical-association.csl #style for references
bibliography: refs.bib #references
link-citations: yes #adds links to the citations
'': default
---

@@

In contrast, generalized additive models (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures. Therefore, GAMs present an excellent choice to analyze non-linear longitudinal data in the context of biomedical research. This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data. We present the basic theory of GAMs, and using reported trends of oxygen saturation in tumors we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate their implementation in R. We also show that GAMs are able to produce estimates that are consistent with the trends of non-linear data even in the case when missing observations exist (with 40% of the simulated observations missing). To make this work reproducible, the code and data used in this paper are available at: https://github.com/aimundo/GAMs-biomedical-research.

### 2021-10-13 Manuscripts/Manuscript by chapters- SIM Revisions/04-Longitudinal analysis with GAMs.Rmd
Started changes in Appendix, need to figure how to break up long code chunks and add text in between.

For the data in Figure \@ref(fig:sim-smooth-plot), A the main effect of interest is how $\mbox{StO}_2$ changes over time for each treatment. To estimate this, the model incorporates independent smooths for _Group_ and _Day_, respectively. The main thing to consider is that model syntax accounts for the fact that one of the variables is numeric (*Day*) and the other is a factor (*Group*). Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the groups. Using $\textsf{R}$ and the package `mgcv` the model syntax is:

```{r, eval = FALSE}
m1 <- gam(StO2_sim ~ Group + s(Day, by=Group, k=5), method='REML', data = dat_sim)
```

This syntax specifies that `m1` will store the model, and that the change in the simulated oxygen saturation (`StO2_sim`) is modeled using independent smooths over `Day` for each `Group` (the parenthesis preceded by `s`) using 5 knots. The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths [@simpson2018]. The parametric term `Group` is added to quantify overall mean differences in the effect of treatment between groups, and the `method` chosen to estimate the smoothing parameters is the restricted maximum likelihood (REML) [@wood2017]. When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of $\mbox{StO}_2$ for each group across time (Figure \@ref(fig:sim-smooth-plot)B). Model diagnostics can be obtained using the `gam.check` function, and the function `appraise` from the package _gratia_ [@gratia]. A guide for model selection and diagnostics is in the [Appendix](#workflow), and an in-depth analysis can be found in [@wood2017] and [@harezlak2018].

@@

m1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),
            method='REML',
            data  = dat_sim)

@@

#adds the predictions to the grid and creates a confidence interval for GAM
gam_predict<-gam_predict%>%
    mutate(fit = predict(m1,gam_predict,se.fit = TRUE,type='response')$fit,
           se.fit = predict(m1, gam_predict,se.fit = TRUE,type='response')$se.fit)

@@

#posthoc comparisons for the linear model
#library(multcomp)

#summary(glht(lm1, linfct = mcp(Group = 'Tukey')))
#summary(glht(lm1, linfct=mcp(Group="Tukey", interaction_average=TRUE)))

@@

Because GAMs do not require equally-spaced or complete observations for all subjects, they are advantageous to analyze longitudinal data where missingness exists. The rationale behind this is that GAMs are able to pick the trend in the data even when some observations are missing. However, this usually causes the resulting smooths to have wider confidence intervals and less ability to pick certain trends. Consider the  simulated $\mbox{StO}_2$ values from Figure \@ref(fig:sim-smooth-plot)B. If 40% of the total observations are randomly deleted and the same interaction GAM fitted for the complete dataset is used, the resulting smooths are still able to show a different trend for each group, but because the empirical Bayesian credible intervals for the smooths overlap during the first 3 days with fewer data points, the trend is less pronounced than in the full dataset (Figure \@ref(fig:sim-smooth-plot)D). Although the confidence intervals have increased for both smooths, the model still shows different trends with as little as 4 observations per group at certain time points.

@@

comp1<-smooth_diff(m1,pdat,'Control','Treatment')

@@

With this expectation of different trends in each group, computing the difference between the trends will identify if the observed change is significant. The difference between groups with similar trends is likely to yield zero, which would indicate that the treatment is not causing a change in the response in one of the groups (assuming the other group is a Control or Reference group).

@@

On the data with missing observations (Figure \@ref(fig:sim-smooth-plot)D), the empirical Bayesian credible intervals of the smooths overlap between days 0 and 3. Consequently, the smooth pairwise comparison (Figure \@ref(fig:plot-pairwise-comp)B) shows that there is no evidence of a significant difference between the groups during that period, but is still able to pick the change on day 3 as the full dataset smooth pairwise comparison.

>>>>

For the data in Figure \@ref(fig:sim-smooth-plot), A the main effect of interest is how $\mbox{StO}_2$ changes over time for each treatment. To estimate this, the model incorporates independent smooths for _Group_ and _Day_, respectively. The main thing to consider is that model syntax accounts for the fact that one of the variables is numeric (*Day*) and the other is a factor (*Group*). Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the group means. Using $\textsf{R}$ and the package `mgcv` the model syntax is:

```{r, eval = FALSE}
gam_02 <- gam(StO2_sim ~ Group + s(Day, by=Group, k=5), method='REML', data = dat_sim)
```

This syntax specifies that `gam_02` (named accordingly so it matches the model sequence from the Appendix) contains the fitted model, and that the change in the simulated oxygen saturation (`StO2_sim`) is modeled using independent smooths over `Day` for each `Group` (the parenthesis preceded by `s`) using 5 knots. The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths [@simpson2018]. The parametric term `Group` is added to quantify overall mean differences in the effect of treatment between groups, and the `method` chosen to estimate the smoothing parameters is the restricted maximum likelihood (REML) [@wood2017]. When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of $\mbox{StO}_2$ for each group across time (Figure \@ref(fig:sim-smooth-plot)B). Model diagnostics can be obtained using the `gam.check` function, and the function `appraise` from the package _gratia_ [@gratia]. A guide for model selection and diagnostics is in the [Appendix](#workflow), and an in-depth analysis can be found in [@wood2017] and [@harezlak2018].

@@

gam_02 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),
            method='REML',
            data  = dat_sim)

@@

#adds the predictions to the grid and creates a confidence interval for GAM
gam_predict<-gam_predict%>%
    mutate(fit = predict(gam_02,gam_predict,se.fit = TRUE,type='response')$fit,
           se.fit = predict(gam_02, gam_predict,se.fit = TRUE,type='response')$se.fit)

@@

@@

Because GAMs do not require equally-spaced or complete observations for all subjects, they are advantageous to analyze longitudinal data where missingness exists. The rationale behind this is that GAMs are able to pick the trend in the data even when some observations are missing. However, this usually causes the resulting smooths to have wider confidence intervals and less ability to pick certain trends. Consider the  simulated $\mbox{StO}_2$ values from Figure \@ref(fig:sim-smooth-plot)B. If 40% of the observations are randomly deleted and the same interaction GAM fitted for the complete dataset is used, the resulting smooths are still able to show a different trend for each group, but because the empirical Bayesian credible intervals for the smooths overlap during the first 3 days with fewer data points, the trend is less pronounced than in the full dataset (Figure \@ref(fig:sim-smooth-plot)D). Although the confidence intervals have increased for both smooths, the model still shows different trends with as few as 4 observations per group at certain time points.

@@

comp1<-smooth_diff(gam_02,pdat,'Control','Treatment')

@@

With this expectation of different trends in each group, computing the difference between the trends will identify if the observed difference is significant. The difference between groups with similar trends is likely to yield zero, which would indicate that the treatment is not causing a change in the response in one of the groups (assuming the other group is a Control or Reference group).

@@

On the data with missing observations (Figure \@ref(fig:sim-smooth-plot)D), the empirical Bayesian credible intervals of the smooths shows the same trend of the full dataset. Consequently, the smooth pairwise comparison (Figure \@ref(fig:plot-pairwise-comp)B) shows that the Control Group has higher $\mbox{StO}_2$ before day 3, and is able to estimate the change on day 3 where the Treatment Group becomes significant as the full dataset smooth pairwise comparison.

### 2021-10-13 Manuscripts/Manuscript by chapters- SIM Revisions/07-Appendix.Rmd
Started changes in Appendix, need to figure how to break up long code chunks and add text in between.

To obtain model diagnostics, two methodologies are to be used: 1) graphical diagnostics, and 2) a model check. In the first case, the functions `appraise` and `draw` from the package _gratia_ can be used to obtain a single output with all the graphical diagnostics. For model check, the functions `gam.check` and `summary` from  _mgcv_ provide detailed information about the model fit and its parameters.

(ref:first-GAM-caption) Graphical diagnostics for the first GAM model. Left: Graphical diagnostics provided by the function `appraise` from the package _gratia_. Right: Fitted smooth for the model, provided by the function `draw`.

@@

#plot(layout2)
```

@@

#### Model check

Special attention must be paid to the 'k-index' from `gam.check`. This parameter indicates if the basis dimension of the smooth is adequate, i.e., it checks that the basis used to create the smooth are adequate to capture the trends in the data. If the model is not adequately capturing the trens in the data, this is indicated by a low k-index value (<1).

@@

gam.check(gam_00)

@@

From the output, it can be seen that the k-index is 0.36, which indicates that the model is not capturing the variability in the data. The `edf` (effective degrees of freedom) is an indicator of the complexity of the smooth. Here the complexity of the smooth is comparable to that of a 4th degree polynomial.

@@

gam.check(gam_01)
summary(gam_01)
```

@@

m1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),
            method='REML',
            data  = dat_sim)

gam.check(m1)
summary(m1)
```

The resulting model is `m1`, which is the model fitted in the main manuscript. By using `appraise()` and `draw` on this model (Figure  \@ref(fig:final-GAM-diag)) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group. From `gam.check`, the k-index is now at an acceptable value ($\approx$ 1.02), and `summary` now indicates that the model is able to capture 89% of the variance in the data.

@@

appr3<-appraise(m1)
sm3<-draw(m1)

@@

AIC(gam_00,gam_01,m1)

@@

The output in this case is expected: model `m1` has a lower AIC (712.46) whereas the initial two models have higher AICs (900 and 858). The AIC should not be considered as the only estimator of model quality, instead to be used as complimentary information to the graphical diagnostics and model checks described above.

@@

The estimation of significant differences between each treatment group can be achieved via pairwise comparisons of the smooth confidence intervals as described in section \@ref(GAM-significance). In this case, the "design matrix" is used to estimate the pairwise comparisons (see main manuscript for details and associated references). Briefly, the "design matrix" (also known as the "Xp matrix") from the selected model (`m1`) is used to calculate a 95% confidence interval of the difference between the smooth terms for each group. This approach allows to estimate the time intervals where a significant difference exists between the groups (confidence interval above or below 0). **All pairwise comparisons in this paper have been centered at the response scale to ease interpretation **.

@@

##matrix that contains the basis functions evaluated at the points in pdat
    xp <- predict(m1, newdata = pdat, type = 'lpmatrix')

@@

    #Multiply matrix by model coefficients. X has (p,n) (rows, columns) and the coefficient matrix has
    #dimensions (n,1). The resulting matrix has dimensions (p,1)
    dif <- X %*% coef(m1)

    #comp<-test %*% coef(gam1)[3:10]

#Calculate standard error for the computed differences using the variance-covariance matrix
    #of the model
    se <- sqrt(rowSums((X %*% vcov(m1, unconditional = FALSE)) * X))
    crit <- qt(0.05/2, df.residual(m1), lower.tail = FALSE)
    #upper  limits
    upr <- dif + (crit * se)
    #lower limits
    lwr <- dif - (crit * se)
    #put all components in a dataframe for plotting
    comp1<-data.frame(pair = paste('Control', 'Treatment', sep = '-'),
               diff = dif,
               se = se,
               upper = upr,
               lower = lwr)

@@

(ref:pairwise-comp-caption) Smooth pairwise comparisons for model `m1` using a 95% confidence interval for the difference between smooths. The comparison is centered at the response scale. Shaded regions indicate time intervals where each treatment group has a non-zero effect.

@@

This section covers the code used to generate Figure \@ref(fig:sim-smooth-plot), where the simulated data, fit of the "final" GAM (`m1`), linear model  and GAM on data with missing observations are presented. Note that panel A in Figure \@ref(fig:sim-smooth-plot) and the inset are generated in the code chunk where the data is simulated in Section \@ref(tumor-data-simulation), and are called later to build the figure.

@@

This code chunk creates panels B and D in Figure \@ref(fig:sim-smooth-plot). Note that this code uses the final GAM from the previous section (`m1`), so the simulated data and the model should be generated before running this section.

@@

#use the function to calculate the difference in smooths
comp1<-smooth_diff(m1,pdat,'Control','Treatment')

>>>>

To obtain model diagnostics, two methodologies are to be used: 1) graphical diagnostics, and 2) a model check. In the first case, the functions `appraise` and `draw` from the package _gratia_ can be used to obtain a single output with all the graphical diagnostics. For model check, the functions `gam.check` and `summary` from  _mgcv_ provide detailed information about the model fit and its parameters. Keep in mind that `gam.check` is a function that also provides the graphical diagnostics obtained using _gratia_, if such graphical output is not desired the source code can be accessed typing `gam.check` in the Console, and the code without the graphical output can be used in a custom function (see \@ref(#gam-00-model-check) for an implementation of this).
(ref:first-GAM-caption) Graphical diagnostics for the first GAM model. Left: Graphical diagnostics provided by the function `appraise` from the package _gratia_. Right: Fitted smooth for the model, provided by the function `draw`.

@@

@@

#### Model check{#gam-00-model-check}

Special attention must be paid to the parameter 'k-index' from `gam.check` (which calls the function `k.check` to perform the calculation). This parameter indicates if the basis dimension of the smooth is adequate, i.e., it checks that the basis used to create the smooth are adequate to capture the trends in the data. If the model is not adequately capturing the trends in the data, this is indicated by a low k-index value (<1). Because we plot the model diagnostics using `appraise` later, the graphical output from `gam.check` will be suppressed by creating a custom function to obtain just the model estimates, thus avoiding repetition of the diagnostic plots. This will be achieved by calling the source code of `gam.check` and using the appropriate code in a new function that will be called `gam.diagnostics`.

@@

#Function that uses the source code of `gam.check` to obtain the estimates without the plots. The source can be checked by typing `gam.check` in the console.
gam.diagnostics<-function (b, old.style = FALSE, type = c("deviance", "pearson", 
    "response"), k.sample = 5000, k.rep = 200, rep = 0, 
    level = 0.9, rl.col = 2, rep.col = "gray80", ...) 
{
    type <- match.arg(type)
    resid <- residuals(b, type = type)
    linpred <- if (is.matrix(b$linear.predictors) && !is.matrix(resid)) 
        napredict(b$na.action, b$linear.predictors[, 1])
    else napredict(b$na.action, b$linear.predictors)

    fv <- if (inherits(b$family, "extended.family")) 
        predict(b, type = "response")
    else fitted(b)
    if (is.matrix(fv) && !is.matrix(b$y)) 
        fv <- fv[, 1]
    gamm <- !(b$method %in% c("GCV", "GACV", "UBRE", 
        "REML", "ML", "P-ML", "P-REML", 
        "fREML"))
    if (gamm) {
        cat("\n'gamm' based fit - care required with interpretation.")
        cat("\nChecks based on working residuals may be misleading.")
    }
    else {
        cat("\nMethod:", b$method, "  Optimizer:", 
            b$optimizer)
        if (!is.null(b$outer.info)) {
            if (b$optimizer[2] %in% c("newton", "bfgs")) {
                boi <- b$outer.info
                cat("\n", boi$conv, " after ", boi$iter, 
                  " iteration", sep = "")
                if (boi$iter == 1) 
                  cat(".")
                else cat("s.")
                cat("\nGradient range [", min(boi$grad), 
                  ",", max(boi$grad), "]", sep = "")
                cat("\n(score ", b$gcv.ubre, " & scale ", 
                  b$sig2, ").", sep = "")
                ev <- eigen(boi$hess)$values
                if (min(ev) > 0) 
                  cat("\nHessian positive definite, ")
                else cat("\n")
                cat("eigenvalue range [", min(ev), ",", 
                  max(ev), "].\n", sep = "")
            }
            else {
                cat("\n")
                print(b$outer.info)
            }
        }
        else {
            if (length(b$sp) == 0) 
                cat("\nModel required no smoothing parameter selection")
            else {
                cat("\nSmoothing parameter selection converged after", 
                  b$mgcv.conv$iter, "iteration")
                if (b$mgcv.conv$iter > 1) 
                  cat("s")
                if (!b$mgcv.conv$fully.converged) 
                  cat(" by steepest\ndescent step failure.\n")
                else cat(".\n")
                cat("The RMS", b$method, "score gradient at convergence was", 
                  b$mgcv.conv$rms.grad, ".\n")
                if (b$mgcv.conv$hess.pos.def) 
                  cat("The Hessian was positive definite.\n")
                else cat("The Hessian was not positive definite.\n")
            }
        }
        if (!is.null(b$rank)) {
            cat("Model rank = ", b$rank, "/", length(b$coefficients), 
                "\n")
        }
    }
    cat("\n")
    kchck <- k.check(b, subsample = k.sample, n.rep = k.rep)
    if (!is.null(kchck)) {
        cat("Basis dimension (k) checking results. Low p-value (k-index<1) may\n")
        cat("indicate that k is too low, especially if edf is close to k'.\n\n")
        printCoefmat(kchck, digits = 3)
    }
}
```

```{r,gam-00-diagnostics, echo=TRUE, include=TRUE}
gam.diagnostics(gam_00)

@@

From the output, it can be seen that the `k-index` is 0.36, which indicates that the model is not capturing the variability in the data. The `edf` (effective degrees of freedom) is an indicator of the complexity of the smooth. Here the complexity of the smooth is comparable to that of a 4th degree polynomial.

@@

gam.diagnostics(gam_01)
summary(gam_01)
```

@@

gam_02 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),
            method='REML',
            data  = dat_sim)

gam.diagnostics(gam_02)
summary(gam_02)
```

The resulting model is `gam_02`, which is the model fitted in the main manuscript. By using `appraise()` and `draw` on this model (Figure  \@ref(fig:final-GAM-diag)) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group. From `gam.check`, the k-index is now at an acceptable value ($\approx$ 1.02), and `summary` now indicates that the model is able to capture 89% of the variance in the data.

@@

appr3<-appraise(gam_02)
sm3<-draw(gam_02)

@@

AIC(gam_00,gam_01,gam_02)

@@

The output in this case is expected: model `gam_02` has a lower AIC (712.46) whereas the initial two models have higher AICs (900 and 858). The AIC should not be considered as the only estimator of model quality, instead to be used as complimentary information to the graphical diagnostics and model checks described above.

@@

The estimation of significant differences between each treatment group can be achieved via pairwise comparisons of the smooth confidence intervals as described in section \@ref(GAM-significance). In this case, the "design matrix" is used to estimate the pairwise comparisons (see main manuscript for details and associated references). Briefly, the "design matrix" (also known as the "Xp matrix") from the selected model (`gam_02`) is used to calculate a 95% confidence interval of the difference between the smooth terms for each group. This approach allows to estimate the time intervals where a significant difference exists between the groups (confidence interval above or below 0). **All pairwise comparisons in this paper have been centered at the response scale to ease interpretation **.

@@

##matrix that contains the basis functions evaluated at the points in pdat
    xp <- predict(gam_02, newdata = pdat, type = 'lpmatrix')

@@

    #Multiply matrix by model coefficients. X has (p,n) (rows, columns) and the coefficient matrix has
    #dimensions (n,1). The resulting matrix has dimensions (p,1)
    dif <- X %*% coef(gam_02)

#Calculate standard error for the computed differences using the variance-covariance matrix
    #of the model
    se <- sqrt(rowSums((X %*% vcov(gam_02, unconditional = FALSE)) * X))
    crit <- qt(0.05/2, df.residual(gam_02), lower.tail = FALSE)
    #upper  limits
    upr <- dif + (crit * se)
    #lower limits
    lwr <- dif - (crit * se)
    #put all components in a dataframe for plotting
    comp1<-data.frame(pair = paste('Control', 'Treatment', sep = '-'),
               diff = dif,
               se = se,
               upper = upr,
               lower = lwr)

@@

(ref:pairwise-comp-caption) Smooth pairwise comparisons for model `gam_02` using a 95% confidence interval for the difference between smooths. The comparison is centered at the response scale. Shaded regions indicate time intervals where each treatment group has a non-zero effect.

@@

This section covers the code used to generate Figure \@ref(fig:sim-smooth-plot), where the simulated data, fit of the "final" GAM (`gam_02`), linear model  and GAM on data with missing observations are presented. Note that panel A in Figure \@ref(fig:sim-smooth-plot) and the inset are generated in the code chunk where the data is simulated in Section \@ref(tumor-data-simulation), and are called later to build the figure.

@@

This code chunk creates panels B and D in Figure \@ref(fig:sim-smooth-plot). Note that this code uses the final GAM from the previous section (`gam_02`), so the simulated data and the model should be generated before running this section.

@@

#use the function to calculate the difference in smooths
comp1<-smooth_diff(gam_02,pdat,'Control','Treatment')
